Haskell in Computational Biology

### 2013-05-25 jfp.tex
line edites sections 1-3

\newcommand\txprobj[3][]{a#1_{{#2}_{j-1}{#3}_j}}
\newcommand\txprobjj[3][]{a#1_{{#2}_{j-1}{#3}_j}}
\newcommand\alignwidth{\ensuremath C} % number of columns in an alignment
\newcommand\pairedwith[1]{{\pi(#1)}}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
\ifpagetuning
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\else
  \ifpagetuning
  And overall,
  \else
  Moreover, measured as a whole, 
  \fi
  the software that incorporates our new
  code 
  performs significantly
  better than the~C++ software that preceded~it.
\fi
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
\ifpagetuning
would be which
\else
experience we were likely to have
\fi
 (\secref{penumbra}).
\item
\emph{[POTENTIAL BULLETS ABOUT SPEED AND ``MORE FUNCTIONAL'']}
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds to the original paper by describing our experience in
trying to simplify and speed up the central algorithm.

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that a homologous protein
has lost an amino acid by deletion from column~$j$.
\hfuzz=1.5pt % fix blot
\end{itemize}
The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.

A hidden Markov model also has distinguished ``begin'' and ``end''
states.
%%
%% for consistency with the text above 'begin' and 'end' are neither
%% capitalized nor italicized.
%%
In~our representation, each state contains a probability or a table of
probabilities, and it is also labeled with one of these labels:
\smallverbatiminput{statelabel}

@@

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
\seclabel{plan7}%
\remark{Reference to \figref{plan7}??}%
\ifpagetuning
``Plan7'' implies that
\else
The model gets its name because
\fi
 there are exactly~7 possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

This model
has begin and end states $B$~and~$E$,
as well as four
nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure 
\ifnotcutting and evolutionary history \fi
with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by~summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
The~path that maximizes the
log of the probability is the most likely path.

@@

\[
\def\maxiquad{\hskip 1.2em\relax}
%\mskip -5mu
\begin{array}{@{}l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
  \vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
  \end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
  \vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
  \end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
  \begin{array}{l@{}c@{}l}
  \vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
  \vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
  \end{array} \right.\\
\end{array}
\]

\caption{Viterbi's original equations}
\figlabel{viterbi-original}

\bigskip

@@

\caption{Viterbi's equations transformed to work with scores}
\figlabel{viterbi-transformed}
\end{figure}

The computation is specified in \figref{viterbi-original}.
A~probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
Probabilities $V_j^I(i)$ and $V_j^D(i)$ are similar.
The
equations are explained very clearly by
\citet[Chapter~5]{Durbin:1998wz}

@@

Our legacy file format also \emph{negates}
the log of each probability, making it a positive number.
%%To~make these numbers more pleasant to read, we negate them,
%%which eliminates hordes of minus signs from our file formats.
The negated logarithm of a probability is called a \emph{score}.
\smallverbatiminput{score}
Type \texttt{Score} has a limited \texttt{Num} instance which permits
scores to be added and subtracted but not multiplied.

@@

\begin{figure}
\def\vsum#1#2{#1+#2}
\[
V_{j}^{\prime \hat s}(i) = e^{\prime}_{\hat s_{j}}(x_{i}) 
  + \min\limits_s \left\{
       \vsum{V_{j-1}^{\prime s}(i - 1)} {a^{\prime}_{s \,\hat
       s}}, \mbox{ where $s$ can precede $\hat s$} \right \}
\]

@@

We also

@@

Here is the general case in the new code:
\smallverbatiminput{hoviterbi}
THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
WE SHOULD PROBABLY FIX THAT.
In the old code, there are three corresponding cases: one each for
match, insert, and delete states.  
Here is the code for the match case:
\smallverbatiminput{viterbi}
The other case are similar; the second argument to \texttt{aScore} is
always equal to the first (state) argument to \texttt{vee'},
and the argument to \texttt{map from} is different in each case.
(In the new code, this argument is identical to \texttt{preceders !
stateRight}.)
NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(

>>>>

\newcommand\txprobj[3][]{a#1_{{#2}_{j-1}{#3}_j}} % probability entering j
\newcommand\txprobxj[3][]{a#1_{{#2}_{j}{#3}_{j+1}}}  % probability exiting j
\newcommand\txprobjj[3][]{a#1_{{#2}_{j}{#3}_j}} % probability within j
\newcommand\alignwidth{\ensuremath C} % number of columns in an alignment
\newcommand\pairedwith[1]{{\pi(#1)}}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell.
Although we encountered some obstacles that kept Haskell from working
as well as we would have liked, our overall experience was positive:
Haskell really worked for us, and we believe it will work equally well
for other computational biologists.
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
and the Haskell code is slower.
But the Haskell code was easy to write,     % coherent subject!
clearly implements the underlying mathematics
 (\secref{viterbi}),
 was easy to parallelize,
and performs well enough
(\secref{perf}).
\ifpagetuning
  And our new tool solves a problem that could not be solved by
  the~C++ tool which preceded~it.
\else
  \ifpagetuning
  And overall,
  \else
  Moreover, measured as a whole, 
  \fi
  the software that incorporates our new
  code 
  performs significantly
  better than the~C++ software that preceded~it.
\fi
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one area:  understanding and
improving
performance (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although the first two authors are computational
biologists
with little functional-programming experience,
Haskell made it easy for us to explore new research ideas.
By~contrast,
our group's C++ code has 
made it hard to explore
% been a good vehicle for exploring 
new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some kept the promise, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
We couldn't tell in advance which 
\ifpagetuning
would be which
\else
experience we were likely to have
\fi
 (\secref{penumbra}).
\item
\emph{[POTENTIAL BULLETS ABOUT SPEED AND ``MORE FUNCTIONAL'']}
\end{itemize}
A~preliminary version of this
paper \citep{daniels:haskell-in-bio:2012} was presented as 
an \emph{Experience Report} in ICFP~2012.
This~paper adds to the original paper by describing our experience in
trying to simplify and speed up the central algorithm.
AND FIX THE CODE AND...

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that a homologous protein
has lost an amino acid by deletion from column~$j$.
A~deletion state contains no tables of probabilities.
\hfuzz=1.5pt % fix blot
\end{itemize}
The probabilities $e_{M_j}(x)$ and $e_{I_j}(x)$ are \emph{emission probabilities}.
The names of the states are easily confused;
one~way to keep them straight is to imagine that the model of the
alignment describes a prototype sequence of length~$\alignwidth$,
and that each state describes an action used to build the query
sequence from the prototype: 
match a prototypical amino~acid,
insert an unexpected amino~acid,
or
delete a prototypical amino~acid.

A~full hidden Markov model also has distinguished ``begin'' and ``end''
states; begin and end states are like match states, but they
correspond to no amino acids and contain no emission probabilities.
%%
%% for consistency with the text above 'begin' and 'end' are neither
%% capitalized nor italicized.
%%
In~our representation, each state is referred to by its position~$j$
within the alignment and by one of these labels:
\smallverbatiminput{statelabel}
We~group states that share a column~$j$; these states are referred
to collectively as ``node~$j$,'' as shown in \figref{plan7}.
This model
has begin and end states $B$~and~$E$,
as well as four
nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.
\remark{We need to figure out if the begin state is really node 0}

@@

The model in \figref{plan7} is from a well-established family
that permits
transitions between match states and all other adjacent states,
but forbids
direct
transitions between insertion 
states and deletion states  \cite{Eddy:1998ut}. 
The family is called the ``Plan7'' family because
every model in the family has 7~transitions per node.
Each transition has its own probability; the probabilities for
transitions leaving states in node~$j$ are
$\txprobjj I I$,
$\txprobxj I M$,
$\txprobjj M I$,
$\txprobxj M M$,
$\txprobxj M D$,
$\txprobxj D M$,
and
$\txprobxj D D$.
When column $j+1$ is a consensus column, the transitions that enter
its match state 
($\txprobxj I M$,
$\txprobxj M M$,
$\txprobxj D M$)
are more likely.
When column $j+1$ is a non-consensus column, the transitions that enter
its insertion and deletion states are more likely.

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure 
with the proteins used to train the model.
The computation finds the most likely matching path through the hidden Markov model.
In~any matching path, each match state and insertion state corresponds to a
unique amino acid in the query sequence.
The probability of the path is the product of all of the corresponding
emission probabilities, together with the product of the transition
probabilities along the edges of the path.

Probabilities are small, and multiplying small numbers makes smaller
numbers.
Using floating-point arithmetic, multiplying probabilities amounts to
storing data in the exponent of a floating-point number, which holds
only 11~bits.
In~order to store data in the significand, which holds 53~bits,
we~instead sum the logarithms of probabilities \cite{Viterbi:1967hq}.
Moreover, in~order to work with positive numbers, 
our legacy file format also \emph{negates}
the logarithm of each probability, making it a positive number.
%%To~make these numbers more pleasant to read, we negate them,
%%which eliminates hordes of minus signs from our file formats.
The negated logarithm of a probability is called a \emph{score}.
\smallverbatiminput{score}
Type \texttt{Score} has a limited \texttt{Num} instance which permits
scores to be added and subtracted but not multiplied.

@@

%%  \[
%%  \def\maxiquad{\hskip 1.2em\relax}
%%  %\mskip -5mu
%%  \begin{array}{@{}l@{}c@{}l}
%%  V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
%%    \begin{array}{l@{}c@{}l}
%%    \vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
%%    \vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
%%    \vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
%%    \end{array} \right.\\[\goo]
%%  V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
%%    \begin{array}{l@{}c@{}l}
%%    \vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
%%    \vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
%%    \end{array} \right.\\[\gum]
%%  V_{j}^{D}(i) &=& \max \left\{
%%    \begin{array}{l@{}c@{}l}
%%    \vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
%%    \vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
%%    \end{array} \right.\\
%%  \end{array}
%%  \]
%%  
%%  \caption{Viterbi's original equations}
%%  \figlabel{viterbi-original}
%%  
%%  
%%  \bigskip

@@

\caption{Viterbi's equations expressed using scores}
\figlabel{viterbi-transformed}
\end{figure}

The~most likely path through a model is the one with the
smallest score.
The~score of each path is computed by dynamic programming; the score associated with
a path to a state~$s$ ($M$, $I$, or~$D$) at node~$j$ and position~$i$ in
the query sequence is written $V_j^{\prime s}(i)$.
For~example, 
score~$V_j^{\prime M}(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
Scores $V_j^{\prime I}(i)$~and~$V_j^{\prime D}(i)$ are similar.
Scores are specified by Viterbi's equations, which are shown
in \figref{viterbi-transformed}. 
The~equations (expressed using probabilities) are explained very clearly by
\citet[Chapter~5]{Durbin:1998wz}

@@

@@

\begin{figure}
\def\vsum#1#2{#1+#2}
\[
V_{j}^{\prime \hat s}(i) = e^{\prime}_{\hat s_{j}}(x_{i}) 
  + \min\limits_s \left\{
       \vsum{V_{\mathit{pnode}\;\hat s\;j}^{\prime
       s}(\mathit{presidue}\;\hat s\;i)} {a^{\prime}_{s \,\hat
       s}}, \mbox{ where $s$ can precede $\hat s$} \right \}
\]

@@

We also combined Viterbi's three equations into one.
Each equation has the same form, but is parameterize over three
values determined by the right-hand state~$\hat s$:
the~index of the previous node ($j$~or~$j-1$),
the~index of the previous residue ($i$~or~$i-1$),
and the set of previous states~$s$ over which the minimum is taken.
If~we abstract the index computations into functions
$\mathit{pnode}$ 
and 
$\mathit{presidue}$
respectively,
we~get the combined, abstract form of Viterbi's equations shown in
\figref{viterbi-abstracted}.
The~implementation of \figref{viterbi-abstracted} looks like this:
\smallverbatiminput{hoviterbi}
Function \texttt{next} makes the recursive call to

@@

Here is the general case in the new code:

THERE IS A GRATUITOUS CHANGE: THE TYPE OF ARRAY USED IS DIFFERENT.
WE SHOULD PROBABLY FIX THAT.
In the old code, there are three corresponding cases: one each for
match, insert, and delete states.  
Here is the code for the match case:
\smallverbatiminput{viterbi}
The other case are similar; the second argument to \texttt{aScore} is
always equal to the first (state) argument to \texttt{vee'},
and the argument to \texttt{map from} is different in each case.
(In the new code, this argument is identical to \texttt{preceders !
stateRight}.)
NORMAN THINKS THIS COMPARISON DOESN'T WORK AT ALL :-(
